---
title: "tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`fibre` is a minimal package for modelling traits along phylogenies using the `INLA` package. We will demonstrate how it works by fitting a model on some simulated data. Let's simulate a simple univariate continuous character that has two distinct rates of evolution on the tree. We first simulate a two-state discrete characater evolving on the tree, then evolve a continuous traits with rates determined by this character. The following code is taken directly from this chapter of Luk J. Harmon's online book (link).

```{r sim_data}
library(fibre)
library(phytools)
library(RRphylo)
library(dplyr)

set.seed(230045)
## transition matrix for the discrete trait simulation
Q <- matrix(c(-1, 1, 1, -1), 2, 2)
## simulated tree
tree <- pbtree(n = 100, scale = 1, d = 0.2)

rates <- setNames(c(1, 20), 1:2)

x <- simulate_traits(tree, rate_model = "discrete", temp_trend_rates = 0,
                     rate_change = Q, rates = rates, internal = TRUE)

# test <- simulate_traits(tree, rate_model = "continuous", temp_trend = -0.5, 
#                         rate_change = 1, internal = TRUE, anc = c(0, 1),
#                         nsim = 1, log_rates = TRUE)

# contMap(tree, test[1:length(tree$tip.label)], anc.states = test[(length(tree$tip.label) + 1L):length(test)], method = "user")


```

Okay, let's fit the most basic model in `fibre` and compare it to what we get with `RRphylo`. As a first step, it is always a good idea to standardise the response variable to improve convergence and comparability.

```{r fit_model}
x <- scale(x) %>% apply(1, function(x) x) ## that last bit converts scaled matrix back to vector

aces_true <- x[(length(tree$tip.label) + 1L):length(x)]

x <- x[1:length(tree$tip.label)]

contMap(tree, x[1:length(tree$tip.label)], anc.states = aces_true, method = "user")


inla_fit <- phybr(phy = tree, data = x, obs_error = "one", verbose = FALSE)
summary(inla_fit)

```
Now for `RRphylo`:

```{r fit_RR}
RR_fit <- RRphylo(tree, x)
n_node <- length(tree$tip.label) + tree$Nnode

plot(cbind(RR_fit$rates[-1], get_rates(inla_fit, "mode")))
```
Let's try fixing the lambda parameter to what `RRphylo` chose.

```{r lambfa_fix}
inla_fit2 <- phybr(phy = tree, data = x, obs_error = "one", hyper = RR_fit$lambda,
                   verbose = FALSE)
summary(inla_fit2)

plot(cbind(RR_fit$rates[-1], get_rates(inla_fit2, "mode")))
abline(0, 1)

inla_fit2 <- phybr(phy = tree, data = x, obs_error = "est", hyper = RR_fit$lambda,
                   verbose = FALSE)
summary(inla_fit2)

plot(cbind(RR_fit$rates[-1], get_rates(inla_fit2, "mode")))
abline(0, 1)
```

```{r full_fit}
inla_fit3 <- phybr(phy = tree, data = x, obs_error = "est",
                   verbose = FALSE)
summary(inla_fit3)

plot(cbind(RR_fit$rates[-1], get_rates(inla_fit3, "mode")))
```
```{r aces_1}
aces_preds <- get_aces(inla_fit3, "mode")
plot(aces_true, aces_preds)

plot(aces_true, RR_fit$aces[ , 1])
```
```{r tips_1}
tip_preds <- get_tips(inla_fit3, "mode")
plot(x, tip_preds)

plot(x, RR_fit$predicted.phenotype[ , 1])

names(tip_preds) <- names(x)
names(aces_preds) <- names(aces_true)

contMap(tree, x[1:length(tree$tip.label)], anc.states = aces_true, method = "user")
contMap(tree, tip_preds, anc.states = aces_preds, method = "user")
contMap(tree, RR_fit$predicted.phenotype %>% apply(1, function(x) x), 
        anc.states = RR_fit$aces %>% apply(1, function(x) x), method = "user")

```

```{r sd_plot}
aces_sd <- get_aces(inla_fit3, "sd")
tip_sd <- get_tips(inla_fit3, "sd")

names(tip_sd) <- names(x)
names(aces_sd) <- names(aces_true)

contMap(tree, tip_sd, anc.states = aces_sd, method = "user")

```

```{r weird_thing}
get_rate_var <- function(RR_fit) {
  betas <- RR_fit$rates
  L <- RR_fit$tip.path
  t <- RR_fit$tree
  Rvar <- array()
          for (i in 1:Ntip(t)) {
              ace.tip <- betas[match(names(which(L[i, ] != 0)), 
                  rownames(betas)), ]
              mat = as.matrix(dist(ace.tip))
              Rvar[i] <- sum(mat[row(mat) == col(mat) + 1])
          }
  var(Rvar)
}
get_rate_var(RR_fit)
```
```{r br_effect}
tree_long <- tree
tree_long$edge.length <- tree_long$edge.length * 111
RR_fit_long <- RRphylo(tree_long, x)
contMap(tree, x[1:length(tree$tip.label)], anc.states = aces_true, method = "user")
contMap(tree, RR_fit_long$predicted.phenotype %>% apply(1, function(x) x), 
        anc.states = RR_fit_long$aces %>% apply(1, function(x) x), method = "user")
RR_fit_long$lambda
get_rate_var(RR_fit_long)
```
```{r opposite}
Q <- matrix(c(-1, 1, 1, -1), 2, 2)
## simulated tree
tree <- pbtree(n = 100, scale = 1)
## simulate discrete character history
tree <- sim.history(tree, Q, anc = "1")
## plot discrete character history on the tree
plotSimmap(tree, setNames(c("blue", "red"), 1:2), pts = F)

x <- sim.rates(tree, setNames(c(20, 1), 1:2), internal = TRUE)

tree$edge.length <- tree$edge.length * 200

x <- scale(x) %>% apply(1, function(x) x) ## that last bit converts scaled matrix back to vector

aces_true <- x[(length(tree$tip.label) + 1L):length(x)]

x <- x[1:length(tree$tip.label)]

contMap(tree, x[1:length(tree$tip.label)], anc.states = aces_true, method = "user")


inla_fit_o <- phybr(phy = tree, data = x, obs_error = "est", verbose = FALSE)
summary(inla_fit_o)

RR_fit_o <- RRphylo(tree, x)
plot(cbind(RR_fit_o$rates[-1], inla_fit_o$summary.fitted.values$mode[200:397]))

ace_ind <- INLA::inla.stack.index(attr(inla_fit_o, "stack"), "aces")$data
ace_preds <- inla_fit_o$summary.fitted.values$mode[ace_ind]
plot(aces_true, ace_preds)
plot(aces_true, RR_fit_o$aces[ , 1])

tip_ind <- 1:length(tree$tip.label)
tip_preds <- inla_fit_o$summary.fitted.values$mode[tip_ind]
plot(x, tip_preds)
plot(x, RR_fit_o$predicted.phenotype[ , 1])

names(tip_preds) <- names(x)
names(ace_preds) <- names(aces_true)

contMap(tree, x[1:length(tree$tip.label)], anc.states = aces_true, method = "user")
contMap(tree, tip_preds, anc.states = ace_preds, method = "user")
contMap(tree, RR_fit_o$predicted.phenotype %>% apply(1, function(x) x), 
        anc.states = RR_fit_o$aces %>% apply(1, function(x) x), method = "user")


```
