---
title: "tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`phinla` is a minimal package for modelling traits along phylogenies using the `INLA` package. We will demonstrate how it works by fitting a model on some simulated data. Let's simulate a simple univariate continuous character that has two distinct rates of evolution on the tree. We first simulate a two-state discrete characater evolving on the tree, then evolve a continuous traits with rates determined by this character. The follwoign code is taken directly from this chapter of Luk J. Harmon's online book (link).

```{r sim_data}
library(phinla)
library(phytools)

set.seed(2300)
## transition matrix for the discrete trait simulation
Q <- matrix(c(-1, 1, 1, -1), 2, 2)
## simulated tree
tree <- pbtree(n = 500, scale = 1)
## simulate discrete character history
tree <- sim.history(tree, Q, anc = "1")
## plot discrete character history on the tree
plotSimmap(tree, setNames(c("blue", "red"), 1:2), pts = F)

x <- sim.rates(tree, setNames(c(1, 20), 1:2), internal = FALSE)
```


```{r setup}

```
